Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2020-01-10 ms/manuscript.Rmd
Use captioner for table and SI captions

---
title: "Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: yes # need for converting tex to docx
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: new-phytologist.sty
header-includes: 
  \usepackage{float}
  \makeatletter\renewcommand*{\fps@figure}{H}\makeatother
editor_options: 
  chunk_output_type: console
nocite: | 
  @Vogel1968
---

@@

# Set up captions
fig_nums <- captioner(prefix = "Fig.")
fig_nums(name = "traits-used-image", caption = "Examples of traits used in this study.")
fig_nums(name = "climate", caption = "Microclimatic variables.")
fig_nums(name = "pca", caption = "Trait PCA.")
fig_nums(name = "tree", caption = "Traits on tree.")
fig_nums(name = "heatmap", caption = "Variable importance scores.")
fig_nums(name = "cwm-div", caption = "Community-weighted means.")
fig_nums(name = "comm-div", caption = "Community diversity metrics.")

# Make short version of citation function for figs
citef <- pryr::partial(fig_nums, display = "cite")

@@

@@

Moorea, French Polynesia (\ang{17;30;}S, \ang{149;50;}W) is a small (\SI{135}{\kilo\meter\squared}) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island (\circa \SI{1.5}{\ma} [@Duncan1976]) has led to a relatively small, yet phylogenetically diverse fern flora (\circa \SI{130}{\species}, 8/11 orders; taxonomy follows @PteridophytePhylogenyGroupI2016; \tab S1).
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations (\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009) and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

@@

We selected several traits relevant to epiphytic growth (\tab \@ref(tab:traits-used) and `r citef("traits-used-image")`).
We preferred morphological traits over physiological traits because data for morphological traits are easier to gather at the scale of our study (\circa \SI{130}{\species} distributed over an elevational gradient from \SIrange{200}{2000}{\meter}).
Plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010].

@@

Morphological traits of gametophytes were all qualitative (either categorical or binary; \tab \@ref(tab:traits-used)).
Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017] or grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (i.e., genus) if documented as a stable taxonomic character.
For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate.
Non-cordate morphotypes include elongate (strap or ribbon \sensu @Farrar2008) and filamentous forms (`r citef("traits-used-image")`).
Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

Trait values and data sources are summarized in Table S2.

@@

Mean, maximum, and standard deviation (SD) of temperature and mean, minimum, and SD of vapor pressure deficit (VPD) were retained after excluding variables with correlation coefficients > 0.9.
Variables related to temperature changed with elevation but did not differ significantly between epiphytic and terrestrial dataloggers (\tab S3; `r citef("climate")`).
Variables related to VPD differed by growth habit and elevation.
Mean and maximum or minimum temperature and VPD all decreased with elevation, whereas SD of temperature increased, but SD of VPD decreased with elevation.
Epiphytic dataloggers had higher values than terrestrial ones for variables related to VPD.

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs \@ref(tab:phylosig-cont), \@ref(tab:phylosig-binary), `r citef("tree")`), and for quantitative traits, different results were obtained for λ and K.
When measured with λ, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of λ close to 1; only number of pinnae pairs had λ close to zero (\tab \@ref(tab:phylosig-cont)).
However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of \K < 1 (\tab \@ref(tab:phylosig-cont)).
All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmae, \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under BM (\tab \@ref(tab:phylosig-binary)).
Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; \tab \@ref(tab:phylosig-binary)).
Similar results were obtained using the "strict" dataset (\tab S4).

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \@ref(tab:pics)).
Epiphytic species tend to have smaller values for all traits (\tab \@ref(tab:pics); contrasts were made by subtracting epiphytic from terrestrial values).
Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, \pval `r stipe_pval`), as well as frond width (\pval `r width_pval`) and degree of laminae dissection (\pval `r dissection_pval`).
For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, \pval `r morphotype_pval`; \tab \@ref(tab:corr-evo)).
Similar results were obtained using the "strict" dataset (\tab S5).

@@

In the full-subsets analysis including all climate variables and diversity metrics, growth habit and temperature emerged as the most important variables (`r citef("heatmap")`).
Most of the best-fitting models included growth habit or the interaction of growth habit and a temperature-related variable (\tab S6).
Spatial autocorrelation was not detected for any of the diversity metrics (\tab S6).

@@

Our study reveals a striking duality in the phylogenetic and functional diversity of fern epiphytes. 
On one hand, we find that fern epiphytes are consistently smaller than terrestrial species at both the species and community level (`r citef("cwm-div")`, \tab \@ref(tab:pics)), and that epiphytic communities are phylogenetically clustered (`r citef("comm-div")`).
At the same time, we find that epiphytic fern communities are significantly more diverse functionally compared to terrestrial communities (`r citef("comm-div")`).
Collectively, these results suggest that environmental filtering plays a strong role in limiting plant size, while simultaneously promoting diversification across other trait axes in epiphytic ferns.

@@

Epiphytes are markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (`r citef("cwm-div")`, \tab \@ref(tab:pics)), supporting both hypotheses (convergence in epiphytic traits and reduced functional diversity of epiphytic communities).
These results are corroborated by @Testo2018a, who similarly observed abrupt size reduction associated with epiphytic growth in a globally sampled fern phylogeny. 
What might explain such consistent and convergent reduction in fern sporophyte size when transitioning into the canopy?
An obvious answer is that water limitation in the canopy exerts strong selective pressure to maintain small size. 
In spite of similar and predictable leaf-level economic tradeoffs between angiosperms and ferns [@Karst2007], fern leaves are limited in their hydraulic flexibility as they rely wholly on primary xylem that feeds into laminar tissues with reduced leaf venation.
Stipe length and structure in particular are critical for regulating whole leaf hydraulic conductance in ferns [@Pittermann2013].
Species growing in water-limited habitats like those in the canopy are expected by biomechanical necessity to decrease stipe length to minimize resistance [@WatkinsJr.2010].
Our results support previous studies that identified a prevalence of short stipes among epiphytic ferns [@WatkinsJr.2010; @Creese2011].
However, given that stipe and frond length, frond width, and rhizome diameter were correlated in our study, and similar traits have previously been shown to co-vary [@Arcand2008; @Creese2011], careful experimental manipulations are required to distinguish the role of each trait in facilitating epiphytism.

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (\tab \@ref(tab:pics)), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (\fig S2).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird's nest ferns (\eg, *Asplenium nidus* L.), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird's nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale have been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

Another trait that defied simple classification into epiphytic or terrestrial groups was specific leaf area (SLA; `r citef("cwm-div")`, \tab \@ref(tab:pics)).
Specific leaf area is part of the leaf economic spectrum (LES), which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a; @Diaz2015].
We expected that epiphytes would exhibit lower SLA, given their more extreme growing conditions.
In contrast, we observed overlapping, wide ranging SLA values in both terrestrial and epiphytic communities, with only slightly lower values in epiphytes (`r citef("cwm-div")`).
This may reflect adaptations to specific niches that are not captured in the binary categorization of habitat commonly applied in ferns. 
For example, the following species ranked among those with the five lowest SLA values (i.e., thick, well-protected leaves) on Moorea: *Humata anderssonii* (Mett.) C.Chr. (a twig epiphyte), *Acrostichum aureum* L. (a terrestrial mangrove specialist), and *Paesia divaticatissima* (Dryand.) Copel. (a terrestrial species occurring on high-elevation ridges) (Table S2).
Each of these species is likely to experience prolonged periods of drought-stress. 
In-depth characterization of such microhabitats was beyond the scope of our study and should be investigated further in the future.

@@

Finally, the presence of hairs and glands in gametophytes also exhibit very high phylogenetic signal (\ie, much greater than expected under Brownian motion; \tab \@ref(tab:phylosig-binary)), and these traits are concentrated in certain taxonomic groups [\eg, hairs, grammitid ferns; glands, Thelypteridaceae; @Stokey1951; @Nayar1971].
Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], and glands may affect rates of osmosis by modifying lipid concentrations [@Crow2011], we did not detect a correlation between hair or gland production and epiphytic growth.
Given their high phylogenetic signal, it is possible that these traits are unrelated to environment but rather reflect phylogenetically conserved developmental pathways [@Johnson2012].
Alternatively, the possibility remains that some epiphytic species rely on hairs or glands to prevent or slow drying depending on their niche within the canopy (\eg, hair-bearing grammitid gametophytes at mid to high elevation).

@@

However, while the canopy clearly acts as a filter, we also find that it simultaneously functions as a promoter of diversity. 
This is evident in our analyses of MPD~func~ and MNTD~func~, which both indicate that epiphytic communities are more functionally diverse compared to terrestrial species (`r citef("comm-div")`), thus contradicting our hypothesis of decreased functional diversity in epiphytic communities.
Additional analyses using different trait combinations support the role of non-size related traits driving this pattern (\fig S1).
We propose that the high functional diversity of epiphytic fern communities is maintained by previously undetected niche partitioning within epiphytic habitats as follows.
The canopy is generally drier than terrestrial habitats, exerting strong selection for small size. 
At the same time, microclimatic conditions in trees vary rapidly over short distances with height and distance from the trunk [@Hietz1995; @Cardelus2005], contributing to greater environmental variation per unit area than terrestrial habitats [@Rojas-Hernandez2018]. 
This vertical and horizontal diversity of niches is occupied by different fern species [@WatkinsJr.2006a; @Cardelus2007; @Parra2009; @Acebey2017].
Thus, when analyzed together, epiphytic fern communities encompass a wider range of morphological variation than terrestrial communities on average.
Assuming that the traits we measured reflect physiological strategies of epiphytes, this high morphological diversity likely derives from physiological adaptations to a variety of within-canopy niches. 
Clearly, fine-grained characterization of canopy niches and associated adaptations of epiphytic ferns occupying them is an important goal for future research.

@@

**Table S1** List of ferns from Moorea, French Polynesia included in this study.

**Table S2** Trait values of ferns on Moorea, French Polynesia.

**Table S3** Analysis of covariance (ANCOVA) for climate variables in response to growth habit (epiphytic vs. terrestrial) with elevation as a covariate.

**Table S4** Phylogenetic signal in binary traits, strict dataset excluding any species whose traits were scored based on taxonomy.

**Table S5** Pagel's (1994) test of correlated evolution between binary traits, strict dataset excluding any species whose traits were scored based on taxonomy.

**Table S6** Best-fitting models explaining community diversity metrics by climate and growth habit.

**Fig. S1** Correlation plot comparing results of functional community diversity analyses using different sets of traits as input.

**Fig. S2** Comparison of community-weighted standard deviations (CWSD) in traits of epiphytic and terrestrial fern communities on Moorea, French Polynesia.

@@

**`r citef("tree")`**: Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Tree obtained from @Nitta2017. Relative value of quantitative (sporophyte) traits shown by heat gradient: low values are lighter, high values are darker. Stipe length and rhizome diameter were log-transformed prior to scaling. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Values for three leaf size traits (stipe length, frond length, and frond with) were correlated, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see \tab \@ref(tab:traits-used). Major epiphytic radiations identified by @Schuettpelz2009 labeled on tree in green: H, Hymenophylloideae; T, Trichomanoideae; V, Vittarioideae; A, *Asplenium*; E, *Elaphoglossum*; P, Polypodiaceae; taxonomy follows @PteridophytePhylogenyGroupI2016.

@@

@@

# Include in PDF using kableExtra to convert to latex format
traits_used %>% 
  rename(`Data type (unit)` = data_type, `Functional significance` = ecological_significance, `Reference(s)` = reference) %>%
  column_to_rownames("trait") %>%
  knitr::kable(caption = "Traits used in this study", format = "latex", booktabs = TRUE, escape = TRUE) %>%
  kable_styling(full_width = FALSE) %>%
  pack_rows("Sporophyte traits", 1, 7) %>%
  pack_rows("Gametophyte traits", 8, 11) %>%
  column_spec(2, width = "8em") %>%
  column_spec(3, width = "12em") %>%
  column_spec(4, width = "8em") %>%
  footnote(general = "Noncordate gametophyte morphotypes include ribbon, strap and filamentous.", general_title = "", threeparttable = TRUE, escape = TRUE)
```

@@

@@

@@

@@

>>>>

---
title: "Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes"
bibliography: "references.bib"
mainfont: Times New Roman.ttf
fontsize: 12pt
mainfontoptions: 
- BoldFont=Times New Roman Bold.ttf
- ItalicFont=Times New Roman Italic.ttf
- BoldItalicFont=Times New Roman Bold Italic.ttf
output: 
  bookdown::pdf_document2:
    citation_package: none # Use pandoc default, otherwise csl can't be used for reference formatting
    toc: no
    number_sections: no
    fig_caption: yes
    keep_tex: yes # need for converting tex to docx
    latex_engine: xelatex
    pandoc_args: [ "--csl", "new-phytologist.csl"]
    includes:
      in_header: new-phytologist.sty
header-includes: 
  - \usepackage{float}
  - \makeatletter\renewcommand*{\fps@figure}{H}\makeatother # Fix position of figs
  - \makeatletter\renewcommand*{\fps@table}{H}\makeatother # Fix position of tables
  - \usepackage{caption} # Needed to turn off automatic numbering of table captions
editor_options: 
  chunk_output_type: console
nocite: | 
  @Vogel1968
---

@@

# Set up captions
# - Figures
fig_nums <- captioner(prefix = "Fig.")
fig_nums(name = "traits-used-image", caption = "Examples of traits used in this study.")
fig_nums(name = "climate", caption = "Microclimatic variables.")
fig_nums(name = "pca", caption = "Trait PCA.")
fig_nums(name = "tree", caption = "Traits on tree.")
fig_nums(name = "heatmap", caption = "Variable importance scores.")
fig_nums(name = "cwm-div", caption = "Community-weighted means.")
fig_nums(name = "comm-div", caption = "Community diversity metrics.")

# - Tables
table_nums <- captioner(prefix = "Table")
table_nums(name = "traits-used", caption = "Traits used in this study")
table_nums(name = "phylosig-cont", caption = "Phylogenetic signal in continuous traits")
table_nums(name = "phylosig-binary", caption = "Phylogenetic signal in binary traits")
table_nums(name = "pics", caption = "Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth")
table_nums(name = "corr-evo", caption = "Pagel's (1994) test of correlated evolution between binary traits")

# - SI figures
s_fig_nums <- captioner(prefix = "Fig. S", auto_space = FALSE, suffix = ": ")
s_fig_nums(name = "corr-plot", "Correlation plot comparing results of functional community diversity analyses using different sets of traits as input.")
s_fig_nums(name = "cwsd", "Comparison of community-weighted standard deviations (CWSD) in traits of epiphytic and terrestrial fern communities on Moorea, French Polynesia.")

# - SI tables
s_table_nums <- captioner(prefix = "Table S", auto_space = FALSE, suffix = ": ")
s_table_nums(name = "species-list", "List of ferns from Moorea, French Polynesia included in this study")
s_table_nums(name = "trait-values", "Trait values of ferns on Moorea, French Polynesia")
s_table_nums(name = "ancova", "Analysis of covariance (ANCOVA) for climate variables in response to growth habit (epiphytic vs. terrestrial) with elevation as a covariate")
s_table_nums(name = "phylosig-binary-strict", "Phylogenetic signal in binary traits, strict dataset excluding any species whose traits were scored based on taxonomy")
s_table_nums(name = "corr-evo-strict", "Pagel's (1994) test of correlated evolution between binary traits, strict dataset excluding any species whose traits were scored based on taxonomy")
s_table_nums(name = "model-fit", "Best-fitting models explaining community diversity metrics by climate and growth habit")

# SI figures

# Make short versions of citation functions
citef <- pryr::partial(fig_nums, display = "cite")
citet <- pryr::partial(table_nums, display = "cite")
citesf <- pryr::partial(s_fig_nums, display = "cite")
citest <- pryr::partial(s_table_nums, display = "cite")

@@

<!-- Turn off automatic numbering of table captions -->
\captionsetup[table]{labelformat=empty}

@@

Moorea, French Polynesia (\ang{17;30;}S, \ang{149;50;}W) is a small (\SI{135}{\kilo\meter\squared}) tropical oceanic island more than \SI{5000}{\kilo\meter} from the nearest continental landmass.
This distance acts as a strong barrier to dispersal [@Carson1995; @Dassler2001], which combined with the young age of the island (\circa \SI{1.5}{\ma} [@Duncan1976]) has led to a relatively small, yet phylogenetically diverse fern flora (\circa \SI{130}{\species}, 8/11 orders; taxonomy follows @PteridophytePhylogenyGroupI2016; `r citest("species-list")`).
Ferns of Moorea include representatives of multiple Cretaceous epiphytic radiations (\eg, *Asplenium*, *Elaphoglossum*, Hymenophylloideae, Polypodiaceae, Vittarioideae; @Schuettpelz2009) and their terrestrial relatives, making them suitable for a comparative study of traits related to the evolution of epiphytic growth in ferns.

@@

We selected several traits relevant to epiphytic growth (`r citet("traits-used")` and `r citef("traits-used-image")`).
We preferred morphological traits over physiological traits because data for morphological traits are easier to gather at the scale of our study (\circa \SI{130}{\species} distributed over an elevational gradient from \SIrange{200}{2000}{\meter}).
Plant morphology and physiology are closely linked, and physiological traits such as rates of photosynthesis and water use efficiency are known to be correlated with morphological traits such as stipe length in ferns [@WatkinsJr.2010].

@@

Morphological traits of gametophytes were all qualitative (either categorical or binary; `r citet("traits-used")`).
Traits were scored by observing gametophytes that were either collected in the field and identified using DNA barcoding [@Nitta2017] or grown from spores of known species in the lab, taken from the literature, or inferred from taxonomic group (i.e., genus) if documented as a stable taxonomic character.
For comparative analysis, gametophyte morphotypes were aggregated into two categories: either cordate (\ie, heart-shaped) or non-cordate.
Non-cordate morphotypes include elongate (strap or ribbon \sensu @Farrar2008) and filamentous forms (`r citef("traits-used-image")`).
Gametophytes were cultivated on Bold's media [@Bold1957] supplemented with Nitch's micronutrients [@Nitsch1951].

@@

Trait values and data sources are summarized in `r citest("trait-values")`.

@@

Mean, maximum, and standard deviation (SD) of temperature and mean, minimum, and SD of vapor pressure deficit (VPD) were retained after excluding variables with correlation coefficients > 0.9.
Variables related to temperature changed with elevation but did not differ significantly between epiphytic and terrestrial dataloggers (`r citest("ancova")`; `r citef("climate")`).
Variables related to VPD differed by growth habit and elevation.
Mean and maximum or minimum temperature and VPD all decreased with elevation, whereas SD of temperature increased, but SD of VPD decreased with elevation.
Epiphytic dataloggers had higher values than terrestrial ones for variables related to VPD.

@@

Most measured traits showed some degree of phylogenetic signal, but the strength varied across traits (\tabs `r table_nums("phylosig-cont", display = "n")`, `r table_nums("phylosig-binary", display = "n")`, `r citef("tree")`), and for quantitative traits, different results were obtained for λ and K.
When measured with λ, most sporophyte traits show phylogenetic signal (as expected under a BM model), with values of λ close to 1; only number of pinnae pairs had λ close to zero (`r citet("phylosig-cont")`).
However, when measured with \K, only frond width and rhizome diameter showed strong phylogenetic signal (more than expected under BM); other traits had values of \K < 1 (`r citet("phylosig-cont")`).
All binary gametophyte traits showed significant phylogenetic signal, either similar to (hairs, \D = `r hairs_d`) or more conserved (gemmae, \D = `r gemmae_d`; glands, \D = `r glands_d`; morphotype, \D = `r morphotype_d`) than expected under BM (`r citet("phylosig-binary")`).
Growth habit showed phylogenetic signal nearly identical to that expected under BM (\D = `r habit_d`; `r citet("phylosig-binary")`).
Similar results were obtained using the "strict" dataset (`r citest("phylosig-binary-strict")`).

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (`r citet("pics")`).
Epiphytic species tend to have smaller values for all traits (`r citet("pics")`; contrasts were made by subtracting epiphytic from terrestrial values).
Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, \pval `r stipe_pval`), as well as frond width (\pval `r width_pval`) and degree of laminae dissection (\pval `r dissection_pval`).
For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, \pval `r morphotype_pval`; `r citet("corr-evo")`).
Similar results were obtained using the "strict" dataset (`r citest("corr-evo-strict")`).

@@

In the full-subsets analysis including all climate variables and diversity metrics, growth habit and temperature emerged as the most important variables (`r citef("heatmap")`).
Most of the best-fitting models included growth habit or the interaction of growth habit and a temperature-related variable (`r citest("model-fit")`).
Spatial autocorrelation was not detected for any of the diversity metrics (`r citest("model-fit")`).

@@

Our study reveals a striking duality in the phylogenetic and functional diversity of fern epiphytes. 
On one hand, we find that fern epiphytes are consistently smaller than terrestrial species at both the species and community level (`r citef("cwm-div")`, `r citet("pics")`), and that epiphytic communities are phylogenetically clustered (`r citef("comm-div")`).
At the same time, we find that epiphytic fern communities are significantly more diverse functionally compared to terrestrial communities (`r citef("comm-div")`).
Collectively, these results suggest that environmental filtering plays a strong role in limiting plant size, while simultaneously promoting diversification across other trait axes in epiphytic ferns.

@@

Epiphytes are markedly smaller than terrestrial species as measured by stipe and frond length, frond width, and rhizome diameter at both the species and community level (`r citef("cwm-div")`, `r citet("pics")`), supporting both hypotheses (convergence in epiphytic traits and reduced functional diversity of epiphytic communities).
These results are corroborated by @Testo2018a, who similarly observed abrupt size reduction associated with epiphytic growth in a globally sampled fern phylogeny. 
What might explain such consistent and convergent reduction in fern sporophyte size when transitioning into the canopy?
An obvious answer is that water limitation in the canopy exerts strong selective pressure to maintain small size. 
In spite of similar and predictable leaf-level economic tradeoffs between angiosperms and ferns [@Karst2007], fern leaves are limited in their hydraulic flexibility as they rely wholly on primary xylem that feeds into laminar tissues with reduced leaf venation.
Stipe length and structure in particular are critical for regulating whole leaf hydraulic conductance in ferns [@Pittermann2013].
Species growing in water-limited habitats like those in the canopy are expected by biomechanical necessity to decrease stipe length to minimize resistance [@WatkinsJr.2010].
Our results support previous studies that identified a prevalence of short stipes among epiphytic ferns [@WatkinsJr.2010; @Creese2011].
However, given that stipe and frond length, frond width, and rhizome diameter were correlated in our study, and similar traits have previously been shown to co-vary [@Arcand2008; @Creese2011], careful experimental manipulations are required to distinguish the role of each trait in facilitating epiphytism.

@@

Despite the apparent constraints on size and the recurrence of non-cordate gametophytes exhibited by epiphytes, we observed greater morphological diversity among epiphytes versus terrestrial species for several traits.
These results do not support the hypothesis that fern epiphytes have reduced functional diversity.
For example, although epiphytes demonstrated less dissected fronds at the species level (`r citet("pics")`), they exhibited much greater variation in this trait relative to terrestrial ferns at the community level (`r citesf("cwsd")`).
It is likely that the high diversity of this trait reflects different physiological adaptations to various canopy niches. 
For instance, filmy ferns (Hymenophyllaceae), which depend on desiccation tolerance to survive dry conditions, have extremely thin, highly dissected fronds.
Bird's nest ferns (\eg, *Asplenium nidus* L.), in contrast, avoid desiccation by capturing detritus and water in the bases of their simple, rosette leaves. 
At low-elevation sites on Moorea where these ferns co-occur, filmy ferns frequently grow in more humid portions of trees including near the main trunk, whereas bird's nest ferns often grow higher in the canopy and on more exposed, drought-prone branches (J. Nitta, pers. obs.). 
Similar predictable patterns of niche partitioning at the local scale have been reported in neotropical epiphtyic ferns as well [@Hietz1998; @Tausz2001]. 
This distribution may reflect alternative water-usage strategies aligned to specific microhabitats in the canopy [desiccation tolerance vs. water capture; @Hietz1998]. 
Such hypotheses need to be carefully tested with additional quantitative investigation, including more finely examining microclimatic variation and species composition across the canopy.

Another trait that defied simple classification into epiphytic or terrestrial groups was specific leaf area (SLA; `r citef("cwm-div")`, `r citet("pics")`).
Specific leaf area is part of the leaf economic spectrum (LES), which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans [@Wright2004a; @Diaz2015].
We expected that epiphytes would exhibit lower SLA, given their more extreme growing conditions.
In contrast, we observed overlapping, wide ranging SLA values in both terrestrial and epiphytic communities, with only slightly lower values in epiphytes (`r citef("cwm-div")`).
This may reflect adaptations to specific niches that are not captured in the binary categorization of habitat commonly applied in ferns. 
For example, the following species ranked among those with the five lowest SLA values (i.e., thick, well-protected leaves) on Moorea: *Humata anderssonii* (Mett.) C.Chr. (a twig epiphyte), *Acrostichum aureum* L. (a terrestrial mangrove specialist), and *Paesia divaticatissima* (Dryand.) Copel. (a terrestrial species occurring on high-elevation ridges) (`r citest("trait-values")`).
Each of these species is likely to experience prolonged periods of drought-stress. 
In-depth characterization of such microhabitats was beyond the scope of our study and should be investigated further in the future.

@@

Finally, the presence of hairs and glands in gametophytes also exhibit very high phylogenetic signal (\ie, much greater than expected under Brownian motion; `r citet("phylosig-binary")`), and these traits are concentrated in certain taxonomic groups [\eg, hairs, grammitid ferns; glands, Thelypteridaceae; @Stokey1951; @Nayar1971].
Although hairs have been hypothesized to reduce rates of drying and increase water uptake after desiccation [@Kappen2007], and glands may affect rates of osmosis by modifying lipid concentrations [@Crow2011], we did not detect a correlation between hair or gland production and epiphytic growth.
Given their high phylogenetic signal, it is possible that these traits are unrelated to environment but rather reflect phylogenetically conserved developmental pathways [@Johnson2012].
Alternatively, the possibility remains that some epiphytic species rely on hairs or glands to prevent or slow drying depending on their niche within the canopy (\eg, hair-bearing grammitid gametophytes at mid to high elevation).

@@

However, while the canopy clearly acts as a filter, we also find that it simultaneously functions as a promoter of diversity. 
This is evident in our analyses of MPD~func~ and MNTD~func~, which both indicate that epiphytic communities are more functionally diverse compared to terrestrial species (`r citef("comm-div")`), thus contradicting our hypothesis of decreased functional diversity in epiphytic communities.
Additional analyses using different trait combinations support the role of non-size related traits driving this pattern (`r citesf("corr-plot")`).
We propose that the high functional diversity of epiphytic fern communities is maintained by previously undetected niche partitioning within epiphytic habitats as follows.
The canopy is generally drier than terrestrial habitats, exerting strong selection for small size. 
At the same time, microclimatic conditions in trees vary rapidly over short distances with height and distance from the trunk [@Hietz1995; @Cardelus2005], contributing to greater environmental variation per unit area than terrestrial habitats [@Rojas-Hernandez2018]. 
This vertical and horizontal diversity of niches is occupied by different fern species [@WatkinsJr.2006a; @Cardelus2007; @Parra2009; @Acebey2017].
Thus, when analyzed together, epiphytic fern communities encompass a wider range of morphological variation than terrestrial communities on average.
Assuming that the traits we measured reflect physiological strategies of epiphytes, this high morphological diversity likely derives from physiological adaptations to a variety of within-canopy niches. 
Clearly, fine-grained characterization of canopy niches and associated adaptations of epiphytic ferns occupying them is an important goal for future research.

@@

**`r citest("species-list")`** List of ferns from Moorea, French Polynesia included in this study.

**`r citest("trait-values")`** Trait values of ferns on Moorea, French Polynesia.

**`r citest("ancova")`** Analysis of covariance (ANCOVA) for climate variables in response to growth habit (epiphytic vs. terrestrial) with elevation as a covariate.

**`r citest("phylosig-binary-strict")`** Phylogenetic signal in binary traits, strict dataset excluding any species whose traits were scored based on taxonomy.

**`r citest("corr-evo-strict")`** Pagel's (1994) test of correlated evolution between binary traits, strict dataset excluding any species whose traits were scored based on taxonomy.

**`r citest("model-fit")`** Best-fitting models explaining community diversity metrics by climate and growth habit.

**`r citesf("corr-plot")`** Correlation plot comparing results of functional community diversity analyses using different sets of traits as input.

**`r citesf("cwsd")`** Comparison of community-weighted standard deviations (CWSD) in traits of epiphytic and terrestrial fern communities on Moorea, French Polynesia.

@@

**`r citef("tree")`**: Time-calibrated phylogenetic tree of ferns from Moorea, French Polynesia with growth habit and associated traits mapped on the tips. Tree obtained from @Nitta2017. Relative value of quantitative (sporophyte) traits shown by heat gradient: low values are lighter, high values are darker. Stipe length and rhizome diameter were log-transformed prior to scaling. States of qualitative (gametophyte) states and growth habit indicated by colors in key. Grey indicates missing data or non-applicable trait states. Values for three leaf size traits (stipe length, frond length, and frond with) were correlated, so of these, we only present stipe length. Species missing data for six or more traits not shown. For a summary of traits, see `r citet("traits-used")`. Major epiphytic radiations identified by @Schuettpelz2009 labeled on tree in green: H, Hymenophylloideae; T, Trichomanoideae; V, Vittarioideae; A, *Asplenium*; E, *Elaphoglossum*; P, Polypodiaceae; taxonomy follows @PteridophytePhylogenyGroupI2016.

@@

**Table `r table_nums("traits-used", display = "n")`** Traits used in this study

@@

# Include in PDF using kableExtra to convert to latex format
traits_used %>% 
  rename(`Data type (unit)` = data_type, `Functional significance` = ecological_significance, `Reference(s)` = reference) %>%
  column_to_rownames("trait") %>%
  knitr::kable(caption = "", format = "latex", booktabs = TRUE, escape = TRUE) %>%
  kable_styling(full_width = FALSE) %>%
  pack_rows("Sporophyte traits", 1, 7) %>%
  pack_rows("Gametophyte traits", 8, 11) %>%
  column_spec(2, width = "8em") %>%
  column_spec(3, width = "12em") %>%
  column_spec(4, width = "8em") %>%
  footnote(general = "Noncordate gametophyte morphotypes include ribbon, strap and filamentous.", general_title = "", threeparttable = TRUE, escape = TRUE)
```

@@

**Table `r table_nums("phylosig-cont", display = "n")`** Phylogenetic signal in continuous traits

@@

**Table `r table_nums("phylosig-binary", display = "n")`** Phylogenetic signal in binary traits

@@

**Table `r table_nums("pics", display = "n")`** Results of phylogenetically independent contrasts analysis of quantitative (sporophyte) traits related to epiphytic growth

@@

**Table `r table_nums("corr-evo", display = "n")`** Pagel's (1994) test of correlated evolution between binary traits
